---
title: "Getting started with aieconindex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with aieconindex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Getting started

`aieconindex` provides tidy R access to the [Anthropic Economic Index](https://www.anthropic.com/economic-index) (AEI) dataset hosted on Hugging Face. The AEI is a recurring release from Anthropic that maps usage of the Claude family of large language models to occupations and tasks using the O*NET taxonomy and the Standard Occupational Classification (SOC) system.

## Installation

```{r}
# install.packages("remotes")
remotes::install_github("charlescoverdale/aieconindex")
library(aieconindex)
```

## Discovering releases

The Anthropic Economic Index is released as dated snapshots (typically every few months). `aei_releases()` lists the snapshots available on Hugging Face.

```{r}
aei_releases()
```

You can fetch the file tree of any single release with `aei_files()`. This is useful when you want to know exactly what is available before downloading anything.

```{r}
aei_files("2025-03-27", recursive = FALSE)
```

## Fetching a usage table

`aei_index()` is the convenience wrapper for the canonical usage table of a release. The shape and exact filename of that table varies across releases (the AEI restructured its directory layout in late 2025); this function papers over that variation.

```{r}
df <- aei_index("latest", source = "claude_ai", variant = "raw")
head(df)
```

The `source` argument selects between Claude.ai consumer traffic (`"claude_ai"`) and first-party API traffic (`"1p_api"`). The `variant` argument selects between raw counts (`"raw"`) and tables already enriched with O*NET and SOC metadata (`"enriched"`); it is ignored for releases from 2026-06-26 onwards, which ship a single file per source.

From the 2026-06-26 release the AEI publishes calendar-month aggregates: one row per date window, geography, category, and metric, with `geo_level`, `category_name`, `hierarchy_level`, `metric_id`, and `node_name` columns. A release can contain more than one month; filter on `date_start` for a single window.

```{r}
df <- aei_index("2026-06-26", source = "claude_ai")
may <- df[df$date_start == max(df$date_start), ]
```

Geographic slices work across schemas. In the monthly schema, subregions carry ISO 3166-2 codes (US states appear as `"US-CA"`):

```{r}
uk <- aei_geography("2026-06-26", country = "GBR")
subs <- aei_geography("2026-06-26", geography = "subregion")
```

Anthropic also publishes standalone labor market impact tables alongside the dated releases:

```{r}
exposure <- aei_labor_market("job_exposure")
penetration <- aei_labor_market("task_penetration")
```

## Fetching arbitrary files

For files that aren't covered by `aei_index()`, use `aei_download()`:

```{r}
soc <- aei_download("2025-03-27", "SOC_Structure.csv")
hierarchy <- aei_download("2025-09-15",
                          "data/output/request_hierarchy_tree_claude_ai.json")
```

CSV files come back as data frames; JSON files come back as parsed lists.

## The aei_tbl class

All data-returning functions emit an object of class `aei_tbl`: a `data.frame` with provenance metadata stored in the `aei_query` attribute. Inspect it directly:

```{r}
attr(df, "aei_query")
```

The class also dispatches a custom `print()`, `summary()`, and `[` method that preserves the metadata when the table is subset.

## Caching

Downloaded files are cached under `tools::R_user_dir("aieconindex", "cache")`. Override with `options(aieconindex.cache_dir = "/your/path")` before the first call. Inspect the cache with `aei_cache_info()` and clear it with `aei_cache_clear()`.

## Citing the data

The Anthropic Economic Index dataset is released under Creative Commons Attribution 4.0 International (CC-BY-4.0). When you use the data in published work, cite it.

```{r}
aei_cite("2026-03-24", format = "bibtex")
```

`aei_cite()` accepts `format = c("text", "bibtex", "bibentry")` and either a release id or `"all"` (the default) to cite the project as a whole.
